How to select predictive models for decision-making or causal inference?

### 2024-10-20 response_to_reviewer.md

*Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
Despite some attempts this is still lacking in literature.  I especially like the simulation scheme which uses different features for data generation and learning models.  Moreover thorough experimental evaluations of causal models are not very frequent making the paper important.  Code is available on Github which is a big plus.*

*I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.*

*Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.*

> We thank the reviewer for the thoughful comments and the call for
> nuance. Our first response is that our thoughts did not come out right
> and that it is not clear, indeed, that R-risk is always the best.
> Rather, that it is a good default solution. We have reformulate parts of
> the paper to convey this nuance better and added the nuance in the
> conclusion.

> In addition, the reviewer is correct that factors beyond overlap
> probably affect which risk is best. We have followed the suggestion to
> explore more diverse simulation scenarii, as detailing the sensitivity
> of our result on the R-risk makes a more nuanced and stronger analysis.
> Thus, we ran a dedicated experiment focusing on the ratio of causal
> effect to background response. We measure the ratio by the absolute
> mean difference between the causal effect and the baseline response,
> measured empirically on each dataset instance $\Delta_{\mu} =
> \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
> \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter
> ($\omega$ in Algorithm 2) in the simulation to force this ratio to
> cover a wide range of values. In 90 dataset instance, it ranges several
> orders of magnitude from 0.005 to 9675. Figure 7 gives the main results
> of this experiment. We see that R-risk is still the best risk on
> average, but when the effect ratio is high (tertile between 9 and
> 9675), the reweighted mu-risk and the the tau-risk_IPW are better than
> the R-risk. However, for low value of the ratio,  the R-risk largely
> outperforms all other metrics. We added these new results to the main
> paper.

### 2024-10-20 response_to_reviewer.md

*I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.*

*Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.*

> We thank the reviewer for the thoughful comments and the call for
> nuance. Our first response is that our thoughts did not come out right
> and that it is not clear, indeed, that R-risk is always the best.
> Rather, that it is a good default solution. We have reformulate parts of
> the paper to convey this nuance better and added the nuance in the
> conclusion.

> In addition, the reviewer is correct that factors beyond overlap
> probably affect which risk is best. We have followed the suggestion to
> explore more diverse simulation scenarii, as detailing the sensitivity
> of our result on the R-risk makes a more nuanced and stronger analysis.
> Thus, we ran a dedicated experiment focusing on the ratio of causal
> effect to background response. We measure the ratio by the absolute
> mean difference between the causal effect and the baseline response,
> measured empirically on each dataset instance $\Delta_{\mu} =
> \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
> \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter
> ($\omega$ in Algorithm 2) in the simulation to force this ratio to
> cover a wide range of values. In 90 dataset instance, it ranges several
> orders of magnitude from 0.005 to 9675. Figure 7 gives the main results
> of this experiment. We see that R-risk is still the best risk on
> average, but when the effect ratio is high (tertile between 9 and
> 9675), the reweighted mu-risk and the tau-risk_IPW are better than
> the R-risk. However, for low value of the ratio,  the R-risk largely
> outperforms all other metrics. We added these new results to the main
> paper.

### 2024-11-10 response_to_reviewer.md

*I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.*

*Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.*

> We thank the reviewer for the thoughful comments and the call for
> nuance. Our first response is that our thoughts did not come out right
> and that it is not clear, indeed, that R-risk is always the best.
> Rather, that it is a good default solution. We have reformulate parts of
> the paper to convey this nuance better and added the nuance in the
> conclusion.

> In addition, the reviewer is correct that factors beyond overlap
> probably affect which risk is best. We have followed the suggestion to
> explore more diverse simulation scenarii, as detailing the sensitivity
> of our result on the R-risk makes a more nuanced and stronger analysis.
> Thus, we ran a dedicated experiment focusing on the ratio of causal
> effect to background response. We measure the ratio by the absolute
> mean difference between the causal effect and the baseline response,
> measured empirically on each dataset instance $\Delta_{\mu} =
> \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
> \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_{0}(x_i) + \mu_{1}(x_i)} \big|$. We vary a dedicated parameter
> ($\omega$ in Algorithm 2) in the simulation to force this ratio to
> cover a wide range of values. In 90 dataset instance, it ranges several
> orders of magnitude from 0.04 to 206. Figure 7 gives the main results
> of this experiment. We see that R-risk is still the best risk on
> average, but when the effect ratio is high (tertile between 9 and
> 9675), the reweighted mu-risk and the tau-risk_IPW are better than
> the R-risk. However, for low value of the ratio,  the R-risk largely
> outperforms all other metrics. We added these new results to the main
> paper. We also studied the effect of the variance of the effect measured as the empirical version of $\frac{var(\mu_1 - \mu_0)}{ var(\mu_1+\mu_0)}$. For this last experiment, we found similar results to the causal effect experiment.

### 2024-11-10 how_to_select_latex.tex
update causal effect xp with the symmetric measure of causal effect.

A given data needs to meet a few assumptions to enable identifying
causal estimands \cite{rubin_causal_2005}. \emph{1)} an individual's
outcome $Y$ is solely governed by the corresponding potential outcome:
\begin{flalign}
    \text{\emph{Consistency} assumption,}
    &&
    Y = A\, Y(1) + (1 - A)\, Y(0)
    \label{eq:consistency}
    &&
\end{flalign}
\emph{2)}
\emph{unconfoundedness} \mbox{$\{Y(0),
        Y(1) \} \indep A | X$}, \emph{~3)} \emph{strong overlap} ie. every patient has a
strictly positive probability to receive each treatment, 
and \emph{4)} \emph{generalization} --no covariate shift.
These classic assumptions, called
\emph{strong ignorability}, are formally detailed in
\ref{apd:causal_assumptions}.

@@

Beyond overlap, we study for caussim simulations, the effect on model selection
of different causal effect ratio to baseline. We vary the empirical mean
absolute difference between the causal effect and the baseline, $\Delta_{\mu} =
    \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)}
    \big|$, covering a ratio range from 0.05 to 1000 ($\text{median}=1.9$, full
distribution in Appendix \ref{apd:effect_ratio}). Figure
\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
better risk.

\begin{figure}[!tb]
    \centering
    \begin{minipage}{\linewidth}
        \includegraphics[width=0.9\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
    \end{minipage}

    \caption{\textbf{R-risk is robust to a wide range of effect ratio}:
    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
    effect ratio are the tertiles of the absolute ratio causal effect to
    baseline response, $\Delta_{\mu}$: Low $[0.005;0.4[$, Medium $[0.44; 9[$,
    High $[9;9675[$. Appendix \ref{apd:effect_ratio} details this
    simulation.}\label{fig:effect_ratio_influence}
\end{figure}

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. 
%
Our conclusions are that without prior knowledge, the R-risk is a good
default. However, there is much remaining variation, and the R-risk will
not be optimal for every situation. We have identified one such specific
situation: when the causal effect is large compared to the variation of
the baseline effect.

@@

Going beyond overlap, we study the effect of another important parameter of the
data generation process: the ratio of causal effect to the baseline response.

For each dataset of our simulation (caussim), we measure this ratio as the absolute mean
difference between the causal effect and the baseline response, measured
empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N
    \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a
dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
ratio to cover a wide range of values. For 179 dataset instances, it ranges
several order of magnitudes with the full distribution of values given in
Table \ref{apd:effect_ratio_distribution}.

\begin{table}[!ht]
    \centering
    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
            \hline
                                   & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
            \textbf{effect\_ratio} & 179            & 159.6         & 852.4        & 0.004674     & 0.004765     & 0.02483       & 0.1566        & 1.879         & 21.12         & 135.1         & 3394          & 9675         \\ \hline
        \end{tabular}
    }
    \caption{Distribution of the causal effect ratio $\Delta_\mu$ for the experiment on causal effect ratio, results in Figure \ref{fig:effect_ratio_influence}.}\label{apd:effect_ratio_distribution}
\end{table}

>>>>

A given data needs to meet a few assumptions to enable identifying
causal estimands \cite{rubin_causal_2005}. \emph{1)} an individual's
outcome $Y$ is solely governed by the corresponding potential outcome:
\begin{flalign}
    \text{\emph{Consistency} assumption,}
     &  &
    Y = A\, Y(1) + (1 - A)\, Y(0)
    \label{eq:consistency}
     &  &
\end{flalign}
\emph{2)}
\emph{unconfoundedness} \mbox{$\{Y(0),
        Y(1) \} \indep A | X$}, \emph{~3)} \emph{strong overlap} ie. every patient has a
strictly positive probability to receive each treatment,
and \emph{4)} \emph{generalization} --no covariate shift.
These classic assumptions, called
\emph{strong ignorability}, are formally detailed in
\ref{apd:causal_assumptions}.

@@

Beyond overlap, we study for caussim simulations, the effect on model selection
of different causal effect ratio to baseline. We vary the empirical mean
absolute difference between the causal effect and the baseline, $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$, covering a ratio range from 0.04 to 206 ($\text{median}=9.1$, full
distribution in Appendix \ref{apd:effect_ratio}). Figure
\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
better risk.

\begin{figure}[!tb]
    \centering
    \begin{minipage}{\linewidth}
        \includegraphics[width=0.9\linewidth]{_6_effect_ratio_sym_by_bin_comparaison_kendall_by_Dataset.pdf}
    \end{minipage}

    \caption{\textbf{R-risk is robust to a wide range of effect ratio}:
    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
    effect ratio are the tertiles of the absolute ratio causal effect to
    baseline response, $\Delta_{\mu}$: Low $[0.04;2.86[$, Medium $[2.86; 16.65[$,
    High $[16.65;206.53[$. Appendix \ref{apd:effect_ratio} details this
    simulation.}\label{fig:effect_ratio_influence}
\end{figure}

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs.
%
Our conclusions are that without prior knowledge, the R-risk is a good
default. However, there is much remaining variation, and the R-risk will
not be optimal for every situation. We have identified one such specific
situation: when the causal effect is large compared to the variation of
the baseline effect, the $\mu\text{-risk}_{IPW}$ performs slightly
better.

@@

Going beyond overlap, we study the effect of another important parameter of the
data generation process: the ratio of causal effect to the baseline response.
Here, we detail the experimental setup of the results presented in Figure \ref{fig:effect_ratio_influence}.

For each dataset of our simulation (caussim), we measure this ratio as the absolute mean
difference between the causal effect and the baseline response, measured
empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$. We vary a
dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
ratio to cover a wide range of values. For 179 dataset instances, it ranges
several order of magnitudes with the full distribution of values given in
Table \ref{apd:effect_ratio_distribution}.

\begin{table}[!ht]
    \centering
    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
            \hline
                                   & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
            \textbf{effect\_ratio} & 179            & 16.95         & 24.92        & 0.043        & 0.054        & 0.32          & 1.34          & 9.12          & 22.55         & 39.97         & 101.12        & 206.53       \\ \hline
        \end{tabular}
    }
    \caption{Distribution of the causal effect ratio $\Delta_\mu$ for the experiment on causal effect ratio, results in Figure \ref{fig:effect_ratio_influence}.}\label{apd:effect_ratio_distribution}
\end{table}
